\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{parskip}

\title{Rust and GitHub Tools -- Questions and Answers}
\author{}
\date{}

\begin{document}

\maketitle

\section*{Question 1}
\textbf{What is the recommended tool for setting up Rust on your system so that it works with Visual Studio Code?}

\begin{itemize}
    \item Codespaces for Rust
    \item GitHub Copilot
    \item \textbf{RustUp the toolchain installer}
    \item Rust Analyzer extension
\end{itemize}

\section*{Question 2}
\textbf{What are some benefits of using dev containers in Codespaces?}

\begin{itemize}
    \item Seamless integration with GitHub Copilot
    \item Automated code suggestions and completions
    \item \textbf{Consistent development environments across different machines}
\end{itemize}

\section*{Question 3}
\textbf{Which tool provides AI pair-programming capabilities?}

\begin{itemize}
    \item RustUp the toolchain installer
    \item Codespaces for Rust
    \item \textbf{GitHub Copilot}
    \item Rust Analyzer extension
\end{itemize}

\section*{Question 4}
\textbf{What is the role of dev containers in Codespaces?}

\begin{itemize}
    \item Customizing the environment in Visual Studio Code
    \item Enhancing code analysis and linting for Rust projects
    \item \textbf{Providing consistent and reproducible development environments}
    \item Enabling AI pair-programming capabilities
\end{itemize}

\section*{Question 5}
\textbf{What are some benefits of using Codespaces for Rust?}

\begin{itemize}
    \item \textbf{Seamless integration with Visual Studio Code}
    \item Streamlined installation process for Rust
    \item \textbf{AI pair-programming capabilities}
    \item Advanced code analysis for Rust projects
\end{itemize}

\section*{Question 6}
\textbf{What is the purpose of using AI pair-programming tools like GitHub Copilot?}

\begin{itemize}
    \item To optimize Codespaces and manage quotas
    \item To provide an introduction to GitHub Copilot
    \item To enable seamless integration with Visual Studio Code
    \item \textbf{To assist developers with code suggestions and completions}
\end{itemize}

\section*{Question 7}
\textbf{What are some advantages of using Codespaces for online development?}

\begin{itemize}
    \item \textbf{Enhanced collaboration through shared development environments}
    \item Automated code analysis and debugging capabilities
    \item Optimized performance for resource-intensive applications
    \item \textbf{Simplified installation process for multiple programming languages}
\end{itemize}

\section*{Question 8}
\textbf{What is the purpose of setting up a Student Account with GitHub?}

\begin{itemize}
    \item Enabling AI pair-programming tools
    \item \textbf{Accessing educational resources and benefits}
    \item Customizing the editor in Visual Studio Code
    \item Understanding usage and quotas of Codespaces
\end{itemize}

\section*{Question 9}
\textbf{What are the benefits of using GitHub Copilot for AI-powered coding assistance?}

\begin{itemize}
    \item Streamlined installation process for multiple programming languages
    \item \textbf{Automated code suggestions and completions}
    \item \textbf{Seamless integration with Visual Studio Code}
    \item Enhanced performance for resource-intensive applications
\end{itemize}

\section*{Question 10}
\textbf{What can developers accomplish using Codespaces?}

\begin{itemize}
    \item Accessing the installation guide for Rust on Windows
    \item Enabling GitHub Copilot for AI-powered coding assistance
    \item Managing quotas and optimizing performance for online development
    \item \textbf{Developing and collaborating on projects online}
\end{itemize}

\end{document}
